---
title: "Monte Carlo Study of the MSE Decomposition in k-NN Regression"
author: "Jiachen Chen, Yiwen Zhao, Yuxi Ren, Jintong Li"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Monte Carlo Study of the MSE Decomposition in k-NN Regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: "bibliography.bib"
nocite: "@*"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
source("../Package Compiling/function5.0.R")
```

## Overview

This vignette demonstrates a minimal Monte Carlo framework to study the **mean squared error (MSE)** decomposition for **k-nearest neighbors (k-NN) regression**. The main purpose is to show how bias, variance, and prediction error behave under a known data-generating process.

We work under a known data generating process (DGP):

\[
Y = f(X) + \varepsilon, \qquad X \sim \mathrm{Unif}(0,1)^d,\quad \varepsilon \sim \mathcal{N}(0,\sigma^2).
\]

The goal is to estimate (pointwise on a grid) the components:

- **Bias**: \( \mathrm{Bias}(x) = E[\hat f(x)] - f(x) \)
- **Variance**: \( \mathrm{Var}(\hat f(x)) \)
- **MSE for estimating the signal**: \( \mathrm{MSE}_f(x) = \mathrm{Bias}(x)^2 + \mathrm{Var}(\hat f(x)) \)
- **MSE for predicting outcomes**: \( \mathrm{MSE}_Y(x) = \mathrm{MSE}_f(x) + \sigma^2 \)

## Main functions used in this vignette

This vignette uses several helper functions from the package. Each function has a different role in the simulation workflow.

- `dgp_x()` generates the predictor matrix \(X\) from a uniform distribution on \([0,1]^d\).
- `dgp_f()` evaluates the true regression function \(f(x)\) under either the 1D or 2D setting.
- `dgp_y()` generates the response variable \(Y\) by adding random noise to the true signal.
- `knn_predict()` computes k-NN regression predictions for a set of test points.
- `make_grid()` creates the evaluation grid on which the decomposition is studied.
- `mc_knn_decomp()` is the main simulation function. It repeats data generation and prediction many times and returns the pointwise bias, variance, and MSE quantities.
- `mc_knn_curve()` computes the average MSE across a set of candidate values of \(k\).
- `plot_mse_1d()` visualizes the pointwise decomposition results in the one-dimensional case.
- `mc_knn_curve_components()` summarizes the grid-averaged bias, variance, and MSE components for each candidate \(k\).
- `plot_curve_components()` plots how these averaged components change across different values of \(k\).

Together, these functions provide a simple simulation framework for studying the bias-variance tradeoff in k-NN regression.

## DGP and target functions

We support two target functions:

1. **Baseline (1D)**: \( f(x) = \sin(2\pi x) \)
2. **Dimensionality extension (2D)**: \( f(x_1, x_2) = \sin(\sqrt{x_1^2 + x_2^2}) \)

```{r}
x <- dgp_x(n = 10, d = 1, seed = 1)
y <- dgp_y(x, sigma = 0.2, type = "1d", seed = 2)
head(cbind(x, y))
```

## Monte Carlo bias–variance decomposition (1D)

Below is a small Monte Carlo run (smaller B for vignette speed; the project default is typically B=500).
Next, we run `mc_knn_decomp()`, which is the main function for the Monte Carlo study. For each repetition, it generates a training sample, applies k-NN regression, predicts on a fixed grid, and then summarizes the pointwise bias, variance, and MSE. In this vignette, we use a smaller value of \(B\) to keep the running time reasonable.
```{r}
res_1d <- mc_knn_decomp(
  n = 200, k = 15, B = 200,
  sigma = 0.2, d = 1, type = "1d",
  grid_size = 101, seed = 123
)

str(res_1d$settings)
head(res_1d$pointwise)
```

## Assumption Check

In our simulation, the predictor variable $X$ is generated independently, and the noise term $\epsilon$ is also generated independently with mean zero and constant variance. This ensures that the observations satisfy the independence and constant variance assumptions required for bias–variance decomposition.

The true regression function is smooth, which makes k-NN regression appropriate for estimating the conditional mean function. Therefore, the assumptions underlying our simulation and analysis are satisfied.

### Visualizing the decomposition

We next visualize the **pointwise** bias–variance decomposition across the input space.
For each grid location \(x\), the Monte Carlo experiment estimates Bias\(^2(x)\), Var\((\hat f(x))\), and the total prediction error \( \text{MSE}_Y(x)\).
As expected, \( \text{MSE}_Y(x) \approx \text{Bias}^2(x) + \text{Var}(\hat f(x)) + \sigma^2 \), and boundary regions may show larger errors due to fewer neighbors on one side.

```{r, fig.width=8, fig.height=5}
plot_mse_1d(res_1d, show = c("bias2", "variance", "mse_y"))
```

### Comparing \hat f(x) (mean prediction) to the true f(x)

This plot overlays the true regression function \(f(x)\) with the Monte Carlo estimate of the mean k-NN predictor \( \mathbb{E}[\hat f(x)] \) at each grid point.
The gap between the two curves visualizes **systematic error (bias)**: larger deviations indicate higher bias for the chosen \(k\).

```{r, fig.width=8, fig.height=5, message=FALSE, warning=FALSE}
library(ggplot2)

pw <- res_1d$pointwise
df <- data.frame(x = pw$x1, f_true = pw$f_true, f_hat_mean = pw$f_hat_mean)

ggplot(df, aes(x = x)) +
  geom_line(aes(y = f_true, color = "True f(x)"), linewidth = 1.2) +
  geom_line(aes(y = f_hat_mean, color = "Monte Carlo mean E[f_hat(x)]"), linewidth = 1.2) +
  scale_color_manual(values = c(
    "True f(x)" = "#0000FF",
    "Monte Carlo mean E[f_hat(x)]" = "#FF00FF"
  )) +
  labs(
    title = "True regression function vs Monte Carlo mean of k-NN predictions",
    x = "x", y = "Value", color = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```


## Scanning k to find an empirical optimal k

We can scan over candidate k values and compute the grid-averaged MSE.

```{r}
k_values <- c(1, 3, 5, 10, 15, 25, 40)
curve <- mc_knn_curve(
  k_values = k_values,
  n = 200, B = 200, sigma = 0.2,
  d = 1, type = "1d",
  grid_size = 101, seed = 10,
  target = "y"
)
curve
```

### Bias–Variance tradeoff across k 

To examine the full bias-variance tradeoff, we use `mc_knn_curve_components()`, which extends `mc_knn_curve()` by keeping the separate averaged components instead of only the total MSE.

For each candidate \(k\), the function:

- runs `mc_knn_decomp()` for \(B\) Monte Carlo repetitions,
- computes pointwise Bias\(^2\), Variance, and MSE on the evaluation grid,
- averages these quantities over the grid.

This output makes it easier to see how increasing \(k\) usually decreases variance but increases bias. We then use `plot_curve_components()` to visualize these averaged components across candidate values of \(k\).

```{r, fig.width=8, fig.height=5}
curve_comp <- mc_knn_curve_components(
  k_values = k_values,
  n = 200, B = 200, sigma = 0.2,
  d = 1, type = "1d",
  grid_size = 101, seed = 10
)

plot_curve_components(curve_comp, show = c("avg_bias2", "avg_variance", "avg_mse_y"))
```

## 2D illustration

The 2D case uses the same functions; predictions are evaluated on a grid on \([0,1]^2\).
For a vignette, use a smaller grid to keep runtime reasonable.

```{r include=FALSE}
res_2d <- mc_knn_decomp(
  n = 200, k = 20, B = 100,
  sigma = 0.2, d = 2, type = "2d",
  grid_size = 31, seed = 99
)
head(res_2d$pointwise)
```


## Test output (example)

In the repository, tests are written in a separate file using **testthat**.  
To run the tests locally:

```r
# In the project root:
# install.packages("testthat")
# testthat::test_file("test.R")
```

(For the Basic Implementation Check submission, include the `test.R` file in the repo.)

## References (BibTeX)

```bibtex
@misc{wickham_r_2023,
	title = {R packages},
	url = {https://r-pkgs.org/},
	urldate = {2026-02-25},
	author = {Wickham, Hadley and Bryan, Jennifer},
	year = {2023},
}

@article{cover_nearest_1967,
	title = {Nearest neighbor pattern classification},
	volume = {13},
	copyright = {https://ieeexplore.ieee.org/Xplorehelp/downloads/license-information/IEEE.html},
	issn = {0018-9448, 1557-9654},
	url = {https://ieeexplore.ieee.org/document/1053964/},
	doi = {10.1109/TIT.1967.1053964},
	number = {1},
	urldate = {2026-02-26},
	journal = {IEEE Transactions on Information Theory},
	author = {Cover, T. and Hart, P.},
	month = jan,
	year = {1967},
	pages = {21--27},
}

@book{robert_monte_2004,
	address = {New York, NY},
	series = {Springer {Texts} in {Statistics}},
	title = {Monte {Carlo} {Statistical} {Methods}},
	copyright = {http://www.springer.com/tdm},
	isbn = {978-1-4419-1939-7 978-1-4757-4145-2},
	url = {http://link.springer.com/10.1007/978-1-4757-4145-2},
	urldate = {2026-02-26},
	publisher = {Springer New York},
	author = {Robert, Christian P. and Casella, George},
	year = {2004},
	doi = {10.1007/978-1-4757-4145-2},
	file = {全文:/Users/renyuxi/Zotero/storage/3LM333I7/Robert和Casella - 2004 - Monte Carlo Statistical Methods.pdf:application/pdf},
}

@book{hastie_elements_2009,
	address = {New York, NY},
	series = {Springer {Series} in {Statistics}},
	title = {The {Elements} of {Statistical} {Learning}},
	copyright = {http://www.springer.com/tdm},
	isbn = {978-0-387-84857-0 978-0-387-84858-7},
	url = {http://link.springer.com/10.1007/978-0-387-84858-7},
	urldate = {2026-02-26},
	publisher = {Springer New York},
	author = {Hastie, Trevor and Tibshirani, Robert and Friedman, Jerome},
	year = {2009},
	doi = {10.1007/978-0-387-84858-7},
}

@book{gareth_james_introduction_2013,
	address = {New York, NY},
	series = {Springer {Texts} in {Statistics}},
	title = {An {Introduction} to {Statistical} {Learning}},
	volume = {103},
	copyright = {http://www.springer.com/tdm},
	isbn = {978-1-4614-7137-0 978-1-4614-7138-7},
	url = {http://link.springer.com/10.1007/978-1-4614-7138-7},
	urldate = {2026-02-26},
	publisher = {Springer New York},
	author = {{Gareth James} and Witten, Daniela and Hastie, Trevor and Tibshirani, Robert},
	year = {2013},
	doi = {10.1007/978-1-4614-7138-7},
}

@misc{openai_chatgpt_2026,
  author = {{OpenAI}},
  title  = {ChatGPT (GPT-5.2 Thinking)},
  year   = {2026},
  url    = {https://chat.openai.com/},
  note   = {Accessed 2026-02-25}
}
```
> **Note:** The figures produced in this R Markdown document are preliminary drafts for illustrating patterns and verifying results. They do not represent the final figures.