---
title: "Monte Carlo Study of the MSE Decomposition in k-NN Regression"
author: "Jiachen Chen, Yiwen Zhao, Yuxi Ren, Jintong Li"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: bibliography.bib
nocite: "@*"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
source("../function R/function.R")
```

## Overview

This vignette demonstrates a minimal Monte Carlo framework to study the **mean squared error (MSE)** decomposition for **k-nearest neighbors (k-NN) regression**.

We work under a known data generating process (DGP):

\[
Y = f(X) + \varepsilon, \qquad X \sim \mathrm{Unif}(0,1)^d,\quad \varepsilon \sim \mathcal{N}(0,\sigma^2).
\]

The goal is to estimate (pointwise on a grid) the components:

- **Bias**: \( \mathrm{Bias}(x) = E[\hat f(x)] - f(x) \)
- **Variance**: \( \mathrm{Var}(\hat f(x)) \)
- **MSE for estimating the signal**: \( \mathrm{MSE}_f(x) = \mathrm{Bias}(x)^2 + \mathrm{Var}(\hat f(x)) \)
- **MSE for predicting outcomes**: \( \mathrm{MSE}_Y(x) = \mathrm{MSE}_f(x) + \sigma^2 \)

## DGP and target functions

We support two target functions:

1. **Baseline (1D)**: \( f(x) = \sin(2\pi x) \)
2. **Dimensionality extension (2D)**: \( f(x_1, x_2) = \sin(\sqrt{x_1^2 + x_2^2}) \)

```{r}
x <- dgp_x(n = 10, d = 1, seed = 1)
y <- dgp_y(x, sigma = 0.2, type = "1d", seed = 2)
head(cbind(x, y))
```

## Monte Carlo bias–variance decomposition (1D)

Below is a small Monte Carlo run (smaller B for vignette speed; the project default is typically B=500).

```{r}
res_1d <- mc_knn_decomp(
  n = 200, k = 15, B = 200,
  sigma = 0.2, d = 1, type = "1d",
  grid_size = 101, seed = 123
)

str(res_1d$settings)
head(res_1d$pointwise)
```

## Assumption Check

In our simulation, the predictor variable $X$ is generated independently, and the noise term $\epsilon$ is also generated independently with mean zero and constant variance. This ensures that the observations satisfy the independence and constant variance assumptions required for bias–variance decomposition.

The true regression function is smooth, which makes k-NN regression appropriate for estimating the conditional mean function. Therefore, the assumptions underlying our simulation and analysis are satisfied.

### Visualizing the decomposition

```{r}
plot_mse_1d(res_1d, show = c("bias2", "variance", "mse_y"))
```

### Comparing \hat f(x) (mean prediction) to the true f(x)

```{r}
pw <- res_1d$pointwise
plot(pw$x1, pw$f_true, type = "l", xlab = "x", ylab = "Value",
     main = "True f(x) vs Monte Carlo mean of k-NN predictions")
lines(pw$x1, pw$f_hat_mean)
legend("topright", legend = c("f(x)", "E[ f_hat(x) ]"), lty = 1, bty = "n")
```


## Scanning k to find an empirical optimal k

We can scan over candidate k values and compute the grid-averaged MSE.

```{r}
k_values <- c(1, 3, 5, 10, 15, 25, 40)
curve <- mc_knn_curve(
  k_values = k_values,
  n = 200, B = 200, sigma = 0.2,
  d = 1, type = "1d",
  grid_size = 101, seed = 10,
  target = "y"
)
curve
```

```{r}
plot(curve$k, curve$avg_mse, type = "b",
     xlab = "k", ylab = "Average MSE over grid",
     main = "k-scan for k-NN (Monte Carlo average MSE)")
```

## 2D illustration

The 2D case uses the same functions; predictions are evaluated on a grid on \([0,1]^2\).
For a vignette, use a smaller grid to keep runtime reasonable.

```{r include=FALSE}
res_2d <- mc_knn_decomp(
  n = 200, k = 20, B = 100,
  sigma = 0.2, d = 2, type = "2d",
  grid_size = 31, seed = 99
)
head(res_2d$pointwise)
```

## Test output (example)

In the repository, tests are written in a separate file using **testthat**.  
To run the tests locally:

```r
# In the project root:
# install.packages("testthat")
# testthat::test_file("test.R")
```

(For the Basic Implementation Check submission, include the `test.R` file in the repo.)

## References (BibTeX)

```bibtex
@misc{wickham_r_2023,
	title = {R packages},
	url = {https://r-pkgs.org/},
	urldate = {2026-02-25},
	author = {Wickham, Hadley and Bryan, Jennifer},
	year = {2023},
}

@article{cover_nearest_1967,
	title = {Nearest neighbor pattern classification},
	volume = {13},
	copyright = {https://ieeexplore.ieee.org/Xplorehelp/downloads/license-information/IEEE.html},
	issn = {0018-9448, 1557-9654},
	url = {https://ieeexplore.ieee.org/document/1053964/},
	doi = {10.1109/TIT.1967.1053964},
	number = {1},
	urldate = {2026-02-26},
	journal = {IEEE Transactions on Information Theory},
	author = {Cover, T. and Hart, P.},
	month = jan,
	year = {1967},
	pages = {21--27},
}

@book{robert_monte_2004,
	address = {New York, NY},
	series = {Springer {Texts} in {Statistics}},
	title = {Monte {Carlo} {Statistical} {Methods}},
	copyright = {http://www.springer.com/tdm},
	isbn = {978-1-4419-1939-7 978-1-4757-4145-2},
	url = {http://link.springer.com/10.1007/978-1-4757-4145-2},
	urldate = {2026-02-26},
	publisher = {Springer New York},
	author = {Robert, Christian P. and Casella, George},
	year = {2004},
	doi = {10.1007/978-1-4757-4145-2},
	file = {全文:/Users/renyuxi/Zotero/storage/3LM333I7/Robert和Casella - 2004 - Monte Carlo Statistical Methods.pdf:application/pdf},
}

@book{hastie_elements_2009,
	address = {New York, NY},
	series = {Springer {Series} in {Statistics}},
	title = {The {Elements} of {Statistical} {Learning}},
	copyright = {http://www.springer.com/tdm},
	isbn = {978-0-387-84857-0 978-0-387-84858-7},
	url = {http://link.springer.com/10.1007/978-0-387-84858-7},
	urldate = {2026-02-26},
	publisher = {Springer New York},
	author = {Hastie, Trevor and Tibshirani, Robert and Friedman, Jerome},
	year = {2009},
	doi = {10.1007/978-0-387-84858-7},
}

@book{gareth_james_introduction_2013,
	address = {New York, NY},
	series = {Springer {Texts} in {Statistics}},
	title = {An {Introduction} to {Statistical} {Learning}},
	volume = {103},
	copyright = {http://www.springer.com/tdm},
	isbn = {978-1-4614-7137-0 978-1-4614-7138-7},
	url = {http://link.springer.com/10.1007/978-1-4614-7138-7},
	urldate = {2026-02-26},
	publisher = {Springer New York},
	author = {{Gareth James} and Witten, Daniela and Hastie, Trevor and Tibshirani, Robert},
	year = {2013},
	doi = {10.1007/978-1-4614-7138-7},
}
```
> **Note:** The figures produced in this R Markdown document are preliminary drafts for illustrating patterns and verifying results. They do not represent the final figures。
