---
title: "Monte Carlo Study of the MSE Decomposition in k-NN Regression"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: bibliography.bib
nocite: "@*"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
devtools::load_all("..")
```

## Overview

This vignette demonstrates a minimal Monte Carlo framework to study the **mean squared error (MSE)** decomposition for **k-nearest neighbors (k-NN) regression**.

We work under a known data generating process (DGP):

\[
Y = f(X) + \varepsilon, \qquad X \sim \mathrm{Unif}(0,1)^d,\quad \varepsilon \sim \mathcal{N}(0,\sigma^2).
\]

The goal is to estimate (pointwise on a grid) the components:

- **Bias**: \( \mathrm{Bias}(x) = E[\hat f(x)] - f(x) \)
- **Variance**: \( \mathrm{Var}(\hat f(x)) \)
- **MSE for estimating the signal**: \( \mathrm{MSE}_f(x) = \mathrm{Bias}(x)^2 + \mathrm{Var}(\hat f(x)) \)
- **MSE for predicting outcomes**: \( \mathrm{MSE}_Y(x) = \mathrm{MSE}_f(x) + \sigma^2 \)

## DGP and target functions

We support two target functions:

1. **Baseline (1D)**: \( f(x) = \sin(2\pi x) \)
2. **Dimensionality extension (2D)**: \( f(x_1, x_2) = \sin(\sqrt{x_1^2 + x_2^2}) \)

```{r}
x <- dgp_x(n = 10, d = 1, seed = 1)
y <- dgp_y(x, sigma = 0.2, type = "1d", seed = 2)
head(cbind(x, y))
```

## Monte Carlo bias–variance decomposition (1D)

Below is a small Monte Carlo run (smaller B for vignette speed; the project default is typically B=500).

```{r}
res_1d <- mc_knn_decomp(
  n = 200, k = 15, B = 200,
  sigma = 0.2, d = 1, type = "1d",
  grid_size = 101, seed = 123
)

str(res_1d$settings)
head(res_1d$pointwise)
```

### Visualizing the decomposition

```{r}
plot_mse_1d(res_1d, show = c("bias2", "variance", "mse_y"))
```

### Comparing \hat f(x) (mean prediction) to the true f(x)

```{r}
pw <- res_1d$pointwise
plot(pw$x1, pw$f_true, type = "l", xlab = "x", ylab = "Value",
     main = "True f(x) vs Monte Carlo mean of k-NN predictions")
lines(pw$x1, pw$f_hat_mean)
legend("topright", legend = c("f(x)", "E[ f_hat(x) ]"), lty = 1, bty = "n")
```

## Scanning k to find an empirical optimal k

We can scan over candidate k values and compute the grid-averaged MSE.

```{r}
k_values <- c(1, 3, 5, 10, 15, 25, 40)
curve <- mc_knn_curve(
  k_values = k_values,
  n = 200, B = 200, sigma = 0.2,
  d = 1, type = "1d",
  grid_size = 101, seed = 10,
  target = "y"
)
curve
```

```{r}
plot(curve$k, curve$avg_mse, type = "b",
     xlab = "k", ylab = "Average MSE over grid",
     main = "k-scan for k-NN (Monte Carlo average MSE)")
```

## 2D illustration

The 2D case uses the same functions; predictions are evaluated on a grid on \([0,1]^2\).
For a vignette, use a smaller grid to keep runtime reasonable.

```{r include=FALSE}
res_2d <- mc_knn_decomp(
  n = 200, k = 20, B = 100,
  sigma = 0.2, d = 2, type = "2d",
  grid_size = 31, seed = 99
)
head(res_2d$pointwise)
```

## Test output (example)

In the repository, tests are written in a separate file using **testthat**.  
To run the tests locally:

```r
# In the project root:
# install.packages("testthat")
# testthat::test_file("test.R")
```

(For the Basic Implementation Check submission, include the `test.R` file in the repo.)

## References (BibTeX)

```bibtex
@book{robert2004monte,
  title={Monte Carlo statistical methods},
  author={Robert, Christian P and Casella, George and Casella, George},
  volume={2},
  year={2004},
  publisher={Springer}
}

@article{cover1967nearest,
  title={Nearest neighbour pattern recognition},
  author={Cover, TM and Hart, PE},
  journal={IEEE Trans. on information theory},
  volume={13},
  number={1},
  pages={21--27},
  year={1967}
}

@article{james2013introduction,
  title={An introduction to statistical learning with applications in R},
  author={James, Gareth and Witten, Daniela and Hastie, Trevor and Tibshirani, Robert},
  year={2013},
  publisher={Springer}
}

@book{hastie2009elements,
  title={The elements of statistical learning: data mining, inference, and prediction},
  author={Hastie, Trevor and Tibshirani, Robert and Friedman, Jerome H and Friedman, Jerome H},
  volume={2},
  year={2009},
  publisher={Springer}
}

@book{wickham2023r,
  title={R packages},
  author={Wickham, Hadley and Bryan, Jennifer},
  year={2023},
  publisher={" O'Reilly Media, Inc."}
}
```
> **Note:** The figures produced in this R Markdown document are preliminary drafts for illustrating patterns and verifying results. They do not represent the final figures。
